Supplemental Information

### 2021-05-05 inst/manuscript/manuscript.Rmd
add metabolite set enrichment analysis panels

  bookdown::word_document2:
    reference_docx: template.docx
    df_print: kable
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      - '--lua-filter=multiple-bibliographies.lua'

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \ref{fig:f1}A), thought to be an indicator of metabolic steady state. Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \ref{fig:f1}D-H). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J, Supplementary Figures 1-3). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program. Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3). In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628]. Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (Supplementary Figure 3), suggesting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}). Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).

@@

To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figures 4-5). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\ref{fig:f3}A-B).

@@

To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \ref{fig:f3A}, Supplementary Figure 6). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \ref{fig:f4}, Supplementary Figure 6, and Supplementary Tables 1-3). Overall, proliferating lung fibroblasts demonstrated high rates of glucose uptake and glycolysis. Approximately 15% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate. In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \ref{fig:f4}A). To facilitate the identification of differential carbon utilization between treatment groups, metabolite fluxes were normalized to cell growth rate. In this analysis, a modest increase in glycolytic flux is observed (Supplementary Figure 4D). This observation suggests that the effects of the HIF-1 transcriptional program are evident only after adjusting for differences in cell growth rate.

Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls. Compared to hypoxia, BAY treatment was associated with an increase in glycolysis and lactate fermentation in 21% oxygen, while similar decreases in serine and glutamine incorporation were observed (Figure \ref{fig:f4}D).

```{r exchange-fluxes}
exch_df <- dplyr::filter(model_fluxes, type == "exch" & id == "MCT")
nm <- c("norm", "hyp", "dmso", "bay")
mct_flux <- dplyr::pull(exch_df, flux) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
mct_lb <- dplyr::pull(exch_df, lb) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
mct_ub <- dplyr::pull(exch_df, ub) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
```

@@

(ref:f3)**Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose. (**B**) Rate values determined from asymptotic regression fit of the data from (A). (**C**) Isotopic labeling of key intracellular metabolites after 72 h of treatment with the indicated tracers. Data are the mean ± SEM of 4 biological replicates. FBP, fructose bisphosphate; 3PG, 3-phosphoglycerate; AKG, α-ketoglutarate.

>>>>

  bookdown::word_document2:
    reference_docx: template.docx
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      - '--lua-filter=multiple-bibliographies.lua'

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \@ref(fig:f1)A), thought to be an indicator of metabolic steady state. Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \@ref(fig:f1)B), but slower growth was not associated with decreased cell viability (Figure \@ref(fig:f1)C). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \@ref(fig:f1)D-H). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \@ref(fig:f1)I) despite activation of the HIF-1 transcriptional program. Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3). In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628]. Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (Supplementary Figure 3), suggesting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \@ref(fig:f2). Similar to hypoxia, BAY decreased cell growth rate (Figure \@ref(fig:f2)A-B) and activated the HIF-1 transcriptional program (Figure \@ref(fig:f2)C-G). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \@ref(fig:f2)H). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \@ref(fig:f2)J).

@@

To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \@ref(fig:f3) and Supplementary Figures 4-5). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\@ref(fig:f3)A-B).

@@

To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \@ref(fig:f3A), Supplementary Figure 6). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \@ref(fig:f4), Supplementary Figure 6, and Supplementary Tables 1-3). Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis. Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (Supplementary Figure 6C). In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \@ref(fig:f4)A) with a significant increase in pentose phosphate pathway flux.

Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of a decreased growth rate. After normalizing metabolite fluxes in normoxia and hypoxia to the associated growth rate, modest increases (~10%) in glycolytic flux were observed (Supplementary Figure 4D). This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program. Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.

By comparison, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (Figure \@ref(fig:f4)B). BAY treatment decreased cell proliferation rate (Figure \@ref(fig:f2)B), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux. Similar to hypoxia, decreases in serine and glutamine incorporation were observed. Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.

## Hypoxia and BAY treatment increase lactate oxidation

Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1 and 2). First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment.

Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively. Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment. This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366]. To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (Figure \@ref(fig:f5), Supplementary Figure 5). Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (Figure \@ref(fig:f5) and Supplementary Figure 4). Interestingly, a similar increase in labeling was not observed in PASMCs (Supplementary Figure 5).

Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high. Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*. In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by `[`^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization. Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718]. In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG). This is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668]. Together, these data suggest that lactate makes a modest (~5% carbon) contribution to this process.

## Hypoxia abrogates the increase in glycolysis following pharmacologic prolyl hydroxylase inhibition

To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (Figure \@ref(fig:f6)A-C). LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen. Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen. Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux. These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.

To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (Figure \@ref(fig:f6)D-F). Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (Figure \@ref(fig:f6)D). These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition. Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (Figure \@ref(fig:f6)E). Several patterns emerged from this analysis (Figure \@ref(fig:f6)F). Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in these proliferating LFs [@RN1603]. The tricarboxylic acid cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine. Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA). A metabolite set enrichment analysis revealed the tricarboxlyic acid cycle to be the most

One of the key differences associated with

@@

(ref:f3) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose. (**B**) Rate values determined from asymptotic regression fit of the data from (A). (**C**) Isotopic labeling of key intracellular metabolites after 72 h of treatment with the indicated tracers. Data are the mean ± SEM of 4 biological replicates. FBP, fructose bisphosphate; 3PG, 3-phosphoglycerate; AKG, α-ketoglutarate.

### 2021-05-05 inst/manuscript/supplement.Rmd
add metabolite set enrichment analysis panels

```{r results = "asis"}
hyp %>% 
  flextable::set_caption("Table 1: Lung fibroblast fluxes in 21% and 0.5% oxygen") %>% 
  flextable::flextable_to_rmd()
```

@@

```{r results = "asis"}
bay %>% 
  flextable::set_caption("Table 2: Lung fibroblast fluxes following DMSO and BAY treatment") %>% 
  flextable::flextable_to_rmd()
```

>>>>

```{r t1, results = "asis"}
hyp %>% 
  flextable::set_caption(
    "Lung fibroblast fluxes in 21% and 0.5% oxygen", 
    autonum = officer::run_autonum()
  ) %>% 
  flextable::flextable_to_rmd()
```

@@

```{r t2, results = "asis"}
bay %>% 
  flextable::set_caption(
    "Lung fibroblast fluxes following DMSO and BAY treatment", 
    autonum = officer::run_autonum()
  ) %>% 
  flextable::flextable_to_rmd()
```
